Viscous Transport in Eroding Porous Media

### 2019-05-11 docs/transport/erosion_transport.tex
doing the transport paper
small changes to notation

The Stokes double-layer potential itself can not represent all solutions
of the incompressible Stokes equations.  Therefore, we complete the
integral equation formulation following the work of Power and
Miranda~\cite{pow-mir1987}.  The completion depends on the Stokeslets
and rotlets
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2} \right), \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, the solution of the Stokes equation is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx),
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0(\eeta](\xx), \quad \xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy) ds_\yy \quad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy) ds_\yy 
    \quad \ell = 1,\ldots,M,
\end{align}
where $\ff$ is the prescribed Dirichlet boundary condition of the
velocity on the outer wall and eroding bodies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
The Stokes double-layer potential~\eqref{eqn:stokesDLP} can be
decomposed as
\begin{align}
  \DDD[\eeta](\xx) &= 
    \frac{1}{2\pi} \int_{\Gamma} 
      \frac{\nn}{\rho^2} (\rr \cdot \eeta) \, ds_\yy + 
    \frac{1}{2\pi} \nabla \int_{\Gamma}
      \frac{\rr \cdot \nn}{\rho^2} (\yy \cdot \eeta) \, ds_\yy \\
    &- \frac{1}{2\pi} x_1 \nabla \int_{\Gamma}
      \frac{\rr \cdot \nn}{\rho^2}\eta_1(\yy) \, ds_\yy -
    \frac{1}{2\pi} x_2 \nabla \int_{\Gamma}
      \frac{\rr \cdot \nn}{\rho^2}\eta_2(\yy) \, ds_\yy.
\end{align}
All these integrals but the first are a gradient of a Laplace layer
potential.  In addition, each component of the first integral can be
written as a Cauchy integral with a complex-valued density
function~\cite{bar-wu-vee2015}.  Therefore, the Stokes double-layer
potential can be written as five Cauchy integrals and their derivatives
with the density functions $\yy \cdot \eeta$, $\eta_1$, and $\eta_2$.
The individual velocity components are
\begin{align}
  \label{eqn:cauchy1}
  u_1(\xx) &= \Re (v[\tau_1](x)) + \Re (v'[\yy\cdot\eeta](x)) 
           -x_1\Re (v'[\eta_1](x)) - x_2\Re (v'[\eta_2](x)), \\
  u_2(\xx) &= \Re (v[\tau_2](x)) - \Im (v'[\yy\cdot\eeta](x)) 
       +x_1\Im (v'[\eta_1](x)) + x_2\Im (v'[\eta_2](x)),
\end{align}
where
\begin{align} 
  \tau_1=(\eta_1+i\eta_2)\frac{\Re(n)}{n}, \quad \text{and} \quad
  \tau_2=(\eta_1+i\eta_2)\frac{\Im(n)}{n},
\end{align}
and $n \in \CC$ is the complex representation of the outward unit normal
$\nn \in \RR^2$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{s:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have shown that the velocity, shear stress, vorticity, and pressure
of the double-layer potential can all be written in terms of Cauchy
integrals and its first two derivatives.  Therefore, we require
quadrature methods to evaluate the expressions in
equation~\eqref{eqn:cauchy}.  Since the integrands are all smooth and
periodic, the Trapezoid rule has spectral accuracy.  However, the
integrand is nearly singular if $x$ is close to $\bd\Omega$, and the
quadrature error is arbitrarily large if the resolution $N$ is fixed.
There have been a variety of near-singular integration schemes
developed, and we extend a method first outlined by Barnett et
al.~\cite{bar-wu-vee2015} which we call the {\em Barycentric quadrature
rule}.  This method first uses the density function $\tau(y)$ to compute
the boundary data of $v$, denoted by $v^-{y}$.  Then, it makes use
of the identity
\begin{align}
  v(x) = \int_{\Gamma} \frac{v^{-}(y)}{y-x} \, dy, 
  \quad x \in \Omega.
\end{align}

Here, we briefly describe this quadrature method, and then derive a
necessary extension to simulate erosion.  The Barycentric quadrature
rule starts with a quadrature rule with nodes $y_j \in \Gamma$ and
weights $w_j$. We use the trapezoid rule because of its superior
accuracy when $x$ is sufficiently far from $\Gamma$.  However, as $x$
approaches $\Gamma$, the integrand becomes nearly singular and the
quality of the trapezoid rule deteriorates. To eliminate the
nearly-singular integrand, we use the identity
\begin{align}
  \int_{\Gamma} \frac{v^{-}(y) - v(x)}{y - x} \,dy = 0.
\end{align}
Then, even if $x$ is close to $\Gamma$, the integrand is no longer
nearly-singular and the accuracy of the trapezoid rule is bounded
independent of $x$.  A similar identity identity involving $v'(x)$ also
holds.   Now, applying the spectrally accurate trapezoid rule, the
quadrature error of the Barycentric quadrature rule
\begin{align}
  v(x) \approx \frac{\sum\limits_{j=1}^N \frac{v_j^-}{y_j-x}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}
  \quad \text{and} \quad
  v'(x) \approx \frac{\sum\limits_{j=1}^N 
    \frac{v_j^- - v(x)}{(y_j - x)^2}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, 
  \label{eqn:vvpQuadrature}
\end{align}
is uniformly bounded~\cite{bar-wu-vee2015}.  A slight modification of
these formulae are applied if $x \in \Omega^{c}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature for $v''(x)$}
While accurate quadrature rules for $v(x)$ and $v'(x)$ have already been
developed~\cite{bar-wu-vee2015}, this work requires a similar quadrature
rule for $v''(x)$.  We mimic the approach developed
in~\cite{bar-wu-vee2015}.  First, if $x \in \Omega$ and $n \in
\mathbb{N}$, then 
\begin{align}
  \frac{1}{2\pi i}\int_{\Gamma} \frac{1}{(y-x)^n}\, dy = 
  \left\{
    \begin{array}{ll}
      1, & n = 1, \\
      0, & n > 1. \\
    \end{array}
  \right.
\end{align}
Applying this identity to the Cauchy integral theorem for $v''(x)$, we
have
\begin{align}
  0 &= \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - v''(x) \\
    &= \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - 
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{(y-x)^2 v''(x)}{(y-x)^3}\,dy -
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v(x)}{(y-x)^3}\, dy -
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2(y-x)v'(x)}{(y-x)^3}\, dy \\
    &= \frac{1}{2\pi i}
      \frac{2v^{-}(y) - v''(x)(y-x)^2 - 2v(x) - 2(y-x)v'(x)}
      {(y-x)^3}\, dy.
\end{align}
Since we have constructed the integrand so that it is bounded for for
all $x \in \Omega$, the quadrature error of the trapezoid rule is
acceptable.  Applying the quadrature rule and solving for $v''(x)$, the
Barycentric quadrature rule for the second derivative is
\begin{align}
  v''(x) &\approx \frac{2\sum\limits_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}.
\end{align}
To compute $v(x)$ and $v'(x)$, the Barycentric quadrature rule in
equation~\eqref{eqn:vvpQuadrature} is used. If $x \in \Omega^c$, the
identity
\begin{align}
\frac{1}{x-a} &= -\frac{1}{2\pi i}\int_{\Gamma} 
    \frac{(y-a)^{-1}}{y-x} dy, 
\end{align}
is used, and the resulting Barycentric quadrature formula is
\begin{align}
  v''(x) &\approx \frac{1}{x-a}\frac{2\sum\limits_{j=1}^N
    \frac{v^{+}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}}
    {\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j}.
\end{align}

@@

To compute the velocity with $\mathcal{O}(N)$ operations, we start by
using the trapezoid rule accelerated with the fast multipole method
(FMM). Since the trapezoid rule is inaccurate when $x$ is close to
$\Gamma$, we replace the trapezoid rule with the Barycentric quadrature
rule for such points.

>>>>

The Stokes double-layer potential itself can not represent all solutions
of the incompressible Stokes equations.  Therefore, we complete the
integral equation formulation following the work of Power and
Miranda~\cite{pow-mir1987}.  The completion depends on the Stokeslets
and rotlets
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2} \right), \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, the solution of the Stokes equation is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx),
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0(\eeta](\xx), \quad \xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy \quad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy 
    \quad \ell = 1,\ldots,M,
\end{align}
where $\ff$ is the prescribed Dirichlet boundary condition of the
velocity on the outer wall and eroding bodies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
The Stokes double-layer potential~\eqref{eqn:stokesDLP} can be
decomposed as
\begin{align}
  \DDD[\eeta](\xx) &= 
    \frac{1}{2\pi} \int_{\bd\Omega} 
      \frac{\nn}{\rho^2} (\rr \cdot \eeta) \, ds_\yy + 
    \frac{1}{2\pi} \nabla \int_{\bd\Omega}
      \frac{\rr \cdot \nn}{\rho^2} (\yy \cdot \eeta) \, ds_\yy \\
    &- \frac{1}{2\pi} x_1 \nabla \int_{\bd\Omega}
      \frac{\rr \cdot \nn}{\rho^2}\eta_1(\yy) \, ds_\yy -
    \frac{1}{2\pi} x_2 \nabla \int_{\bd\Omega}
      \frac{\rr \cdot \nn}{\rho^2}\eta_2(\yy) \, ds_\yy.
\end{align}
All these integrals but the first are a gradient of a Laplace layer
potential.  In addition, each component of the first integral can be
written as a Cauchy integral with a complex-valued density
function~\cite{bar-wu-vee2015}.  Therefore, the Stokes double-layer
potential can be written as five Cauchy integrals and their derivatives
with the density functions $\yy \cdot \eeta$, $\eta_1$, and $\eta_2$.
The individual velocity components are
\begin{align}
  \label{eqn:cauchy1}
  u_1(\xx) &= \Re (v[\tau_1](x)) + \Re (v'[\yy\cdot\eeta](x)) 
           -x_1\Re (v'[\eta_1](x)) - x_2\Re (v'[\eta_2](x)), \\
  u_2(\xx) &= \Re (v[\tau_2](x)) - \Im (v'[\yy\cdot\eeta](x)) 
       +x_1\Im (v'[\eta_1](x)) + x_2\Im (v'[\eta_2](x)),
\end{align}
where
\begin{align} 
  \tau_1=(\eta_1+i\eta_2)\frac{\Re(n)}{n}, \quad \text{and} \quad
  \tau_2=(\eta_1+i\eta_2)\frac{\Im(n)}{n},
\end{align}
and $n \in \CC$ is the complex representation of the outward unit normal
$\nn \in \RR^2$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{s:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have shown that the velocity, shear stress, vorticity, and pressure
of the double-layer potential can all be written in terms of Cauchy
integrals and its first two derivatives.  Therefore, we require
quadrature methods to evaluate the expressions in
equation~\eqref{eqn:cauchy}.  Since the integrands are all smooth and
periodic, the Trapezoid rule has spectral accuracy.  However, the
integrand is nearly singular if $x$ is close to $\bd\Omega$, and the
quadrature error is arbitrarily large if the resolution $N$ is fixed.
There have been a variety of near-singular integration schemes
developed, and we extend a method first outlined by Barnett et
al.~\cite{bar-wu-vee2015} which we call the {\em Barycentric quadrature
rule}.  This method first uses the density function $\tau(y)$ to compute
the boundary data of $v$, denoted by $v^-{y}$.  Then, it makes use
of the identity
\begin{align}
  v(x) = \int_{\bd\Omega} \frac{v^{-}(y)}{y-x} \, dy, 
  \quad x \in \Omega.
\end{align}

Here, we briefly describe this quadrature method, and then derive a
necessary extension to simulate erosion.  The Barycentric quadrature
rule starts with a quadrature rule with nodes $y_j \in \bd\Omega$ and
weights $w_j$. We use the trapezoid rule because of its superior
accuracy when $x$ is sufficiently far from $\bd\Omega$.  However, as $x$
approaches $\bd\Omega$, the integrand becomes nearly singular and the
quality of the trapezoid rule deteriorates. To eliminate the
nearly-singular integrand, we use the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^{-}(y) - v(x)}{y - x} \,dy = 0.
\end{align}
Then, even if $x$ is close to $\bd\Omega$, the integrand is no longer
nearly-singular and the accuracy of the trapezoid rule is bounded
independent of $x$.  A similar identity identity involving $v'(x)$ also
holds.   Now, applying the spectrally accurate trapezoid rule, the
quadrature error of the Barycentric quadrature rule
\begin{align}
  v(x) \approx \frac{\sum\limits_{j=1}^N \frac{v_j^-}{y_j-x}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}
  \quad \text{and} \quad
  v'(x) \approx \frac{\sum\limits_{j=1}^N 
    \frac{v_j^- - v(x)}{(y_j - x)^2}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, 
  \label{eqn:vvpQuadrature}
\end{align}
is uniformly bounded~\cite{bar-wu-vee2015}.  A slight modification of
these formulae are applied if $x \in \Omega^{c}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature for $v''(x)$}
While accurate quadrature rules for $v(x)$ and $v'(x)$ have already been
developed~\cite{bar-wu-vee2015}, this work requires a similar quadrature
rule for $v''(x)$.  We mimic the approach developed
in~\cite{bar-wu-vee2015}.  First, if $x \in \Omega$ and $n \in
\mathbb{N}$, then 
\begin{align}
  \frac{1}{2\pi i}\int_{\bd\Omega} \frac{1}{(y-x)^n}\, dy = 
  \left\{
    \begin{array}{ll}
      1, & n = 1, \\
      0, & n > 1. \\
    \end{array}
  \right.
\end{align}
Applying this identity to the Cauchy integral theorem for $v''(x)$, we
have
\begin{align}
  0 &= \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - v''(x) \\
    &= \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - 
    \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{(y-x)^2 v''(x)}{(y-x)^3}\,dy -
    \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{2v(x)}{(y-x)^3}\, dy -
    \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{2(y-x)v'(x)}{(y-x)^3}\, dy \\
    &= \frac{1}{2\pi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y) - v''(x)(y-x)^2 - 2v(x) - 2(y-x)v'(x)}
      {(y-x)^3}\, dy.
\end{align}
Since we have constructed the integrand so that it is bounded for for
all $x \in \Omega$, the quadrature error of the trapezoid rule is
acceptable.  Applying the quadrature rule and solving for $v''(x)$, the
Barycentric quadrature rule for the second derivative is
\begin{align}
  v''(x) &\approx \frac{2\sum\limits_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}.
\end{align}
To compute $v(x)$ and $v'(x)$, the Barycentric quadrature rule in
equation~\eqref{eqn:vvpQuadrature} is used. If $x \in \Omega^c$, the
identity
\begin{align}
\frac{1}{x-a} &= -\frac{1}{2\pi i}\int_{\bd\Omega} 
    \frac{(y-a)^{-1}}{y-x} dy, 
\end{align}
is used, and the resulting Barycentric quadrature formula is
\begin{align}
  v''(x) &\approx \frac{1}{x-a}\frac{2\sum\limits_{j=1}^N
    \frac{v^{+}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}}
    {\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j}.
\end{align}

@@

To compute the velocity with $\mathcal{O}(N)$ operations, we start by
using the trapezoid rule accelerated with the fast multipole method
(FMM). Since the trapezoid rule is inaccurate when $x$ is close to
$\bd\Omega$, we replace the trapezoid rule with the Barycentric
quadrature rule for such points.
